---
title: "R. palustris- E. coli interaction Model with purine transfer"
authors: "James B. McKinlay"
date: "`r format(Sys.time(), '%d %B, %Y')`"
header-includes:
   - \usepackage{array}
output: pdf_document
geometry: margin=2.54cm
---

## Overview
Monod model simulating population and extracellular metabolism in a coculture of an E.coli purine auxotroph and R. palustris. Under anaerobic conditions, E. coli ferments glucose to excreted products, some of which serve as a carbon source for R. palustris. R. palustris excretes adenine which is required by the E. coli auxotroph. The following attempts to captures these basic features using a set of coupled differential equations.
To simplify the output, the model does not include fermentation products that do not contribute to acidification or cross-feeding, namely ethanol, H2, and CO2.

## Set working directory
```{r}
rm(list=ls()) # clears your working directory
getwd() # tells you your current working directory
setwd("/Users/jmckinla/OneDrive - Indiana University/Lab_files_personal/Manuscripts/Purine coculture/Purine excretion mechanism") # sets your working directory
```

## Install packages
```{r}
#install.packages("deSolve") # run this the first time by removing the ‘#’
require("deSolve") # installs any dependencies
```

## Mutualism function
```{r}
mutfun <- function(times, y, parms) {
  with(as.list(c(y, parms)), { 
  
   uEc <- uEcMax*(G/(Kg+G))*(P/(Kp+P))*(bEc/(bEc+10^(f+C)))
 
    # E.coli growth rate (h^-1); controlled by Monod function where G = glucose concentration, Kg = half saturation constant for G, P = purine (adenine) concentration, Kp = half saturation constant for P; bEc/(bEc+10^(f+C)) is a sigmoidal function resembling a buffer titration curve where accumulation of formate (f) and consumable organic acids (C) will decrease the growth rate
 
    uRp <- uRpMax*(C/(Kc+C))*(bRp/(bRp+10^(f+C))) 

    # R. palustris growth rate (h^-1); controlled by Monod function where C = consumable organic acid (OAcs) concentration, Kc = half saturation constant for C; growth rate is also controlled by acid accumulation as described for the E. coli growth rate

  dGdt <- -uEc*Ec/Yg   -Ec*(G/(Kg+G))*(10/(10+1.09^(1000*uEc)))*(bEc/(bEc+10^(f+C)))*((100/(100+6^C))*(rc+rf)) 

# change in glucose (mM/h); the 2nd function describes growth independent metabolism as described by McCully et al, 2017, Environ Microbiol; applying the Monod and sigmoidal acid accumulation functions constrains growth independent metabolism to be dependent on glucose availability and permissive acid levels

  dPdt <-  -uEc*Ec/Yp + Rp*(IP-P)*Perm*SA*(C/(Kc+C))*(bRp/(bRp+10^(f+C))) 

   # change in purine (adenine) conc (mM/h); Ec = E. coli cells/ml; Yp = growth yield on adenine; Rp = Rpalustris cells/ml; IP = intracellular adenine conc (umol/cm^3); P = extracellular adenine conc (umol/cm^3); Perm = adenine permeability coefficient (cm/s); SA = cellular surface area (cm^2/cell) ; applying the Monod and sigmoidal acid accumulation functions constrains diffusion to be dependent on organic acid availability and permissive acid levels

   dCdt <- -uRp*Rp/Yc  + uEc*Ec*Fc + Ec*(G/(Kg+G))*(10/(10+1.09^(1000*uEc)))*(bEc/(bEc+10^(f+C)))*(rc*(100/(100+6^C)))#+rc_mono) 

# change in organic acids (OAcs; mM/h); ; Yc = Rp growth yield on OAcs; Fc = OAcs excretion level (umol/Ec cell); the 2nd function describes growth independent metabolism as described by McCully et al, 2017, Environ Microbiol;
   
   dFdt <- uEc*Ec*Ff + Ec*(G/(Kg+G))*(10/(10+1.09^(1000*uEc)))*(bEc/(bEc+10^(f+C)))*(rf*(100/(100+6^C)))# + rf_mono)
   
# change in formate (mM/h); Ff = formate excretion (umol/Ec cell); the 2nd function describes growth independent metabolism as described by McCully et al, 2017, Environ Microbiol;

    dRpdt <-Rp*uRp 
    
#change in Rp cell density (cells/ml)

    dEcdt <- Ec*uEc

 #change in Ec cell density (cells/ml)

    return(list(c(dGdt, dPdt, dCdt, dFdt, dRpdt, dEcdt))) 
  })
}
```

## Parameters and initial conditions
```{r}

# Parameters
parms <- c(

#### E.coli ####  

uEcMax = 0.28, # E. coli max growth rate (h^-1; McCully et al, 2017, Environ Microbiol)

Yg = 4.4*10^6,# E. coli cells/umol; based on time course data in Fig 2

Yp = 6*10^9, # E. coli cells/umol adenine; based on bioassay std curve in Fig 4 

Fc = 3.44*10^-7, # umol organic acids / E. coli cell; Based on Fig 2: the sum of organic acids left over in coculture (2.20*10^-7 umol/ E.coli cell) + what was estimated to be assimilated by R. palustris (1.24*10^-7) 

Ff = 2*10^-7, # umol/cell; based on what is left over in coculture

Kg = 0.02, # E. coli half saturation constant for glucose (mM; Buhr et al 1992 J Biol Chem)

Kp = 0.004, # E. coli half saturation constant for adenine (mM; 0.004 = average for PurP (0.001) and YicO (0.0065); Papakostas et al. 2013. JBC 288: P36827-36840 

rc = 300*10^-11, # mM glucose converted to organic acids per E. coli cell/ml per h in the absence of growth (umol/cell/h; McCully et al, 2017, Environ Microbiol)

rf = 50*10^-11,# mM glucose converted to formate per E. coli cell/ml/h in the absence of growth (umol/cell/h; rounded up from McCully et al, 2017, Environ Microbiol)

#### R. palustris #####
uRpMax = 0.0875, #R. palustris max growth rate with NH4Cl (h^-1 )

Yc = 2*10^8, # R. palustris growth yield on organic acids (cells/umol OAcs consumed; based on values from this study)

Kc = 0.01, # half saturation constant for organic acids (mM; assumed)

Perm = 1.38e-5, # adenine permeability coefficient (cm/s; Xiang and Anderson, 1994, J Membrane Biol, 140, 111-122)

SA = 5.1e-8, # R. palustris surface area (5.1e-8 cm^2; based on a 3.25 um cell length; LaSarre 2018 mBio, Fig 3F) Smaller and larger SA values were upper and lower bounds as follows: 4.24e-8 cm^2 for 2.7 um cell length; 5.97e-8 cm^2 for 3.8 um cell length

IP = 1.5, # Intracellular adenine concentration (mM or umol/cm3; 1.5 mM for CGA009; 0.02 mM for TIE-1)

bEc = 1*10^43, # resistance of E. coli to acid; reflects medium buffering (McCully et al, 2017, Environ Microbiol)

bRp = 1*10^32) # resistance of R. palustris to acid; reflects medium buffering (McCully et al, 2017, Environ Microbiol)

# initial concentrations:
init <- c( 
         G = 22.5, # Glucose (mM)  
         P = 0.00000001, # adenine (mM); non-zero to initiate growth 
         C = 0.00000001, # OAcs (mM); non-zero to initiate growth; set to 10 for Rp monocultures
         f = 0, # Formate (mM)
         Rp = 3000000, #(cells/ml; based on Fig 2B)
         Ec = 6000000) #(cells/ml; based on Fig 2B); set to 0 for Rp monocultures

# Time range and time step
Tmax = 200 # hours
TimeStep = 1 # hours
Time = seq(0, Tmax, by = TimeStep)
```

## Output
```{r}
out <-  ode(y = init, times = Time, func = mutfun, parms = parms, method = "lsode")
#head(out) # prints first 6 time points
#(out) #prints all time points
tail(out) # prints last 6 time points
write.csv(out, file = "out.csv") # creates comma-delimited file that can be pasted into spreadsheet for graphing
```

## Plotting
```{r}

plot(out[,1], out[,7], xlab = "time (hr)", ylab = "Ec", type = "l", las = 1, 
     col = "orange", lwd = 4, ylim = c(0, 1.1*max(out[,7]))) # plot P2

par(new=T)
plot(out[,1], out[,6], xlab="", ylab ="", type = "l",las = 1,
     col = "purple", ylim=c(0,1*max(out[,6])) )
axis(4, ylim=c(0,1.2*max(out[,6])), lwd=2, line=3.5)
mtext(4, text="Rpal", line=5.5) # plot R. palustris biomass

par(new=T)
plot(out[,1], out[,4], xlab="", ylab ="",  type = "l",las = 1,
     col = "blue", ylim=c(0,1.3*max(out[,4])) )
axis(4, ylim=c(0,1.2*max(out[,4])), lwd=2, line=3.5)
mtext(4, text="Oacs", line=5.5) # plot organic acids

par(new=T)
plot(out[,1], out[,5], xlab="", ylab ="",  type = "l",las = 1,
   col = "green", ylim=c(0,1.3*max(out[,5])) )
axis(4, ylim=c(0,0.1*max(out[,5])), lwd=2, line=3.5)
mtext(4, text="Formate", line=5.5) # plot formate

par(new=T)
plot(out[,1], out[,3], xlab="", ylab ="",  type = "l",las = 1,
     col = "red", ylim=c(0,1*max(out[,3])) )
axis(4, ylim=c(0,1*max(out[,3])), lwd=4, line=3.5)
mtext(4, text="purine", line=5.5) # plot purine
par(new=T)

plot(out[,1], out[,2], xlab="", ylab ="",  type = "l",las = 1,
     col = "black", ylim=c(0,1*max(out[,2])) )
axis(4, ylim=c(0,1*max(out[,2])), lwd=3, line=3.5)
mtext(4, text="glucose", line=5.5) # plot glc
```
