---
title: "R. palustris- E. coli interaction Model with N2 limitation"
authors: "James B. McKinlay"
date: "`r format(Sys.time(), '%d %B, %Y')`"
header-includes:
   - \usepackage{array}
output: pdf_document
geometry: margin=2.54cm
---

## Overview
Monod model that explores interactions between Esherichia coli and Rhodopseudomonas palustris. Under anaerobic conditions, glucose is consumed by E. coli and converts some of this to fermentation byproducts, which serve as a carbon source for R. palustris. In turn, R. palustris fixes atmospheric nitrogen gas, which is converted into ammonium and serves as E. coli's sole source of nitrogen. The following attempts to captures these basic features using a set of coupled differential equations.
This model was used in: http://biorxiv.org/content/early/2016/10/25/083386
This model differs from that presented in LaSarre et al. 2016. ISME J. in the following ways:

(i) a sigmoidal function, rather than a Monod function, is used control the transition to growth-independent fermentation by E. coli; (10/(10+1.09^(1000*uEc)))

(ii) sigmoidal functions are used to transition from NH4+ excretion (1-(40/(40+1.29^N))) to H2 production (40/(40+1.29^N)) by R. palustris as N2 becomes limiting 

(iii) a sigmoidal function is used to simulate the inhibiting effects of accumulated organic acids on not only growth, but now also metabolism, for both species; (bx/(bx+10^(f+C)))

(iv) a sigmoidal function is used to dampen growth-independent product formation rates when consumable organic acids accumulate, thereby simulating the slow growth-independent fermentation rates observed in E. coli monocultures where organic acids accumulate, compared to faster fermentation rates in cocultures where organic acids are consumed (r*(100/(100+6^C)) + r_mono); in this case, rmono is the monoculture growth-independent fermentation rate, whereas r is an upper value observed in coculture; the feedback term uses ^C instead of ^(C+f), otherwise the function incorrectly halts formate production in cocultures even when C is fully consumed

(v) H2 production by R. palustris is coupled to the depletion of consumable organic acids, assuming that 0.5 CO2 are produced per H2 (McKinlay et al. 2014. JBC. 289:1960-1970)

(vi) the R. palustris Km value for N2 (Kn) was given a value of 6, based on the change in growth rate at limiting N2 concentrations in coculture 

(vii) product formation parameter (R and r) values were increased to more accurately simulate observed growth rates in coculture

(viii) the E. coli acid resistance parameter (bEc) was increased relative to that for R. palustris (bRp)

## Set working directory
```{r}
rm(list=ls()) # clears your working directory
getwd() # tells you your current working directory
setwd("~/GitHub/Task3") # sets your working directory
```

## Install packages
```{r}
#install.packages("deSolve") # run this first time
require("deSolve") # installs any dependencies
```

## Mutualism function
```{r}
mutfun <- function(times, y, parms) {
  with(as.list(c(y, parms)), { 
    uEc <- uEcMax*(G/(Kg+G))*(A/(Ka+A))*(bEc/(bEc+10^(f+C))) # E. coli growth rate
    uRp <- uRpMax*(N/(Kn+N))*(C/(Kc+C))*(bRp/(bRp+10^(f+C))) # R. palustris growth rate
    dGdt <- -uEc*Ec/Yg - (uEc*Ec*(Rc+Rf+Re+Rco2)) - Ec*(G/(Kg+G))*(10/(10+1.09^(1000*uEc)))*(bEc/(bEc+10^(f+C)))*((100/(100+6^C))*(rc+rf+re+rco2)+rc_mono+rf_mono+re_mono+rco2_mono)# change in glucose (mM/h); 
    dNdt <- -(uRp*Rp*0.5*Ra)*(1-(40/(40+1.29^N))) -(uRp*Rp/Yn) # change in N2 (mM N2/h; 0.5 accounts for half an N2 consumed for every NH4+ made)
    dCdt <- Ec*uEc*Rc*2 + Ec*2*(G/(Kg+G))*(10/(10+1.09^(1000*uEc)))*(bEc/(bEc+10^(f+C)))*(rc*(100/(100+6^C))+rc_mono) - (uRp*Rp/Yc) - 0.25*Rp*uRp*RhRp - 0.25*Rp*rHp*(C/(Kc+C))*(40/(40+1.29^N))*(bRp/(bRp+10^(f+C))) # change in organic acids except formate (mM/h); 0.25 takes into account both electrons sent to H2 and accomanying carbon to CO2
    dFdt <- (Ec*uEc*Rf*6) + Ec*6*(G/(Kg+G))*(10/(10+1.09^(1000*uEc)))*(bEc/(bEc+10^(f+C)))*(rf*(100/(100+6^C)) + rf_mono) # change in formate (mM/h);  
    dAdt <- (Rp*uRp*Ra)*(1-(40/(40+1.29^N))) - (uEc*Ec/Ya) # change in NH4+ (mM/h) 
    dEcdt <- uEc*Ec  # change in E. coli (cells/ml/h)
    dRpdt <- uRp*Rp   # R. palustris (cells/ml/h)
    dedt <- Ec*uEc*Re*3 + Ec*3*(G/(Kg+G))*(10/(10+1.09^(1000*uEc)))*(bEc/(bEc+10^(f+C)))*(re*(100/(100+6^C))+re_mono) # change in ethanol (mM/h)
    dCO2dt <- (Ec*uEc*Rco2*6) + Ec*6*(G/(Kg+G))*(10/(10+1.09^(1000*uEc)))*(bEc/(bEc+10^(f+C)))*(rco2*(100/(100+6^C))+rco2_mono) + Rp*uRp*RhRp*0.5 + Rp*rHp*0.5*(C/(Kc+C))*(40/(40+1.29^N))*(bRp/(bRp+10^(f+C))) #change in CO2 (mM/h; concentration assumes all CO2 is dissolved); 0.5 CO2 made per H2 based on Rpal resting cell values from McKinlay et al. 2014. JBC. 289:1960-1970.
    dHdt <- Rp*uRp*RhRp + Rp*rHp*(C/(Kc+C))*(40/(40+1.29^N))*(bRp/(bRp+10^(f+C)))+ Ec*uEc*RhEc + Ec*(G/(Kg+G))*(10/(10+1.09^(1000*uEc)))*(bEc/(bEc+10^(f+C)))*(rH*(100/(100+6^C))+rH_mono) #change in H2 (mM/h; concentration assumes all H2 is dissolved) 
    return(list(c(dGdt, dNdt, dCdt, dFdt, dAdt, dEcdt, dRpdt, dedt, dCO2dt, dHdt))) 
  })
}
```

## Parameters and initial conditions
```{r}
# Parameters
parms <- c(
           uEcMax = 0.28, # E. coli max growth rate (h^-1)
           uRpMax = 0.0772, # R. palustrix max growth rate (h^-1)
           Yg = 8*10^7, # Ec growth yield (cells/umol glucose consumed)
           Yn = 5*10^8, # R. palustris growth yield (cells/umol N2 consumed; assumes all N2 is dissolved)
           Yc = 2.5*10^8, #R. palustris growth yield on organic acids (cells/umol Oacs consumed) 
           Ya = 1*10^9, #E. coli growth yield on NH4+ (cells/umol NH4+ consumed)
           Kg = 0.02, # E. coli half saturation constant for glucose (mM; Buhr et al 1992 J Biol Chem 267:3847-51)
           Kc = 0.01, # R. palustris half saturation constatnt for organic acids (mM; assumed)
           Kn = 6, # R. palustris half saturation constant for N2; based on fitting a MM curve against uRp vs N2 plot
           Ka = 0.01, # E. coli half saturation constant for NH4+ (mM; Khademi et al. 2004 Science 305:1587-1594)
           Rc = 1.9*10^-8, # mM glucose converted to succinate, acetate, and lactate per E. coli cell/ml (umol/cell)
           Rf = 8*10^-9,# mM glucose converted to formate per E. coli cell/ml (umol/cell) 
           Re = 4.5*10^-9,# mM glucose converted to EtOH per E. coli cell/ml (umol/cell) 
           Rco2= 5*10^-10, # mM glucose converted to CO2 per E. coli cell/ml (umol/cell) 
           Ra = 0.15*10^-9, # mM N2 converted to NH4+ per R. palustris Nx cell/ml (umol/Nx cell)
           RhRp = 2*10^-9, # mM H2 produced per R. palustris cell/ml; (umol/cell)
           RhEc = 5*10^-9, # mM H2 produced per E. coli cell/ml; (umol/cell)
           rc = 300*10^-11, # mM glucose converted to organic acids per E. coli cell/ml per h in the absence of growth (umol/cell/h)
          rc_mono = 1.2*10^-11, # basal rate of mM glucose converted to organic acids per E. coli cell/ml/h in the absence of growth, e.g., in E.coli monoculture (umol/cell/h); divided by 2 from previous model since not being doubled by Rpal pull anymore; scaled up by 1.1 to account for new sigmoidal function that only achieves 0.9 x a max value
           rf = 47*10^-11,# mM glucose converted to formate per E. coli cell/ml/h in the absence of growth (umol/cell/h)
          rf_mono = 0.83*10^-11, # basal rate of mM glucose converted to formate per E. coli cell/ml/h in the absence of growth, e.g., in E.coli monoculture (umol/cell/h)
           re = 15*10^-11, # mM glucose converted to ethanol per E. coli cell/ml per h in the absence of growth (umol/cell/h)
          re_mono = 0.5*10^-11, # basal rate of mM glucose converted to ethanol per E. coli cell/ml/h in the absence of growth, e.g., in E.coli monoculture (umol/cell/h)
           rco2 = 2*10^-11, # mM glucose converted to CO2 per E. coli cell/ml/h in the absence of growth (umol/cell/h)
          rco2_mono = 1.3*10^-11, # basal rate of mM glucose converted to CO2 per E. coli cell/ml/h in the absence of growth, e.g., in E.coli monoculture (umol/cell/h)
           rH = 2*10^-11, # mM H2/E. coli cell/ml/h in the absence of growth (umol/cell/h)
          rH_mono = 0.83*10^-11, # basal rate of mM H2 production per E. coli cell/ml per h in the absence of growth, e.g., in E.coli monoculture (umol/cell/h)
           rHp = 27*10^-11,# mM H2/R. palustris cell/ml/h in the absence of growth (umol/cell/h)
           bEc = 1*10^43, # resistance of E. coli to acid; reflects medium buffering
           bRp = 1*10^32) #resistance of R. palustris to acid; reflects medium buffering
# initial concentrations below
init <- c( 
          G = 25, # Glucose (mM)  
          N = 70, # N2 (mM) 
          C = 0, # Consumable organic acids (mM)
          f = 0, # Formate (mM)
          A = 0.00005, # NH4+ (mM) (default value reflects amount from ammonium molybdate)
          Ec = 4000000, # E. coli (cells/ml)
          Rp = 36000000, # R. palustris (cells/ml)
          e = 0, #ethanol (mM)
          CO2 = 0, #CO2 (mM)
          H = 0) #H2 (mM)
          # Time range and time step
Tmax = 300 # hours
TimeStep = 1 #hours
Time = seq(0, Tmax, by = TimeStep)
```

## Output
```{r}
out <-  ode(y = init, times = Time, func = mutfun, parms = parms, method = "lsode")
head(out) # prints first 6 time points
tail(out)
#out# prints all time points
write.csv(out, file = "out.csv") # creates comma-delimited file that can be pasted into spreadsheet for graphing
```

## Plotting
```{r}
plot(out[,1], out[,7], xlab = "time (hr)", ylab = "Ecoli", type = "l", las = 1, 
     col = "gold", lwd = 4, ylim = c(0, 1*max(out[,7]))) # plot E. coli biomass

par(new=T)
plot(out[,1], out[,8], xlab="", ylab ="", type = "l",las = 1,
     col = "purple", ylim=c(0,1.2*max(out[,8])) )
axis(4, ylim=c(0,1.2*max(out[,7])), lwd=2, line=3.5)
mtext(4, text="Rpal", line=5.5) # plot R. palustris biomass

par(new=T)
plot(out[,1], out[,4], xlab="", ylab ="",  type = "l",las = 1,
     col = "green", ylim=c(0,1.3*max(out[,4])) )
axis(4, ylim=c(0,1.2*max(out[,4])), lwd=2, line=3.5)
mtext(4, text="Oacs", line=5.5) # plot organic acids

par(new=T)
plot(out[,1], out[,6], xlab="", ylab ="",  type = "l",las = 1,
   col = "blue", ylim=c(0,1.3*max(out[,6])) )
axis(4, ylim=c(0,0.1*max(out[,4])), lwd=2, line=3.5)
mtext(4, text="NH4+", line=5.5) # plot ammonium

par(new=T)
plot(out[,1], out[,2], xlab="", ylab ="",  type = "l",las = 1,
     col = "black", ylim=c(0,1.4*max(out[,2])) )
axis(4, ylim=c(0,1.2*max(out[,2])), lwd=2, line=3.5)
mtext(4, text="glucose", line=5.5) # plot glucose

par(new=T)
plot(out[,1], out[,5], xlab="", ylab ="",  type = "l",las = 1,
     col = "gray", ylim=c(0,1.2*max(out[,5])) )
axis(4, ylim=c(0,1.2*max(out[,5])), lwd=2, line=3.5)
mtext(4, text="formate", line=5.5) # plot formate

par(new=T)
plot(out[,1], out[,10], xlab="", ylab ="",  type = "l",las = 1,
     col = "red", ylim=c(0,1*max(out[,10])) )
axis(4, ylim=c(0,1.2*max(out[,10])), lwd=2, line=3.5)
mtext(4, text="H2", line=5.5) # plot hydrogen
```
